PAPER DRAFT

### 2021-11-04 Papers/Polymethylation_scores/02_abstract.Rmd
more revisions post Kelly 1st round comments

**Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.  DNA methylation signatures vary by prenatal maternal smoke exposure in blood and placental samples and can function as a biomarker. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues. Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking associates with salivary DNA methylation summary measures, including global methylation, epigenetic age, and polymethylation scores. We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.

**Results**: We created a polymethylation score using coefficients from a metanalysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. Prenatal maternal smoking was associated with this polymethylation score () but not epigenetic age or global methylation (). The association with the polymethylation score was consistent at age 9 () and age 15 ().  The direction and magnitude of the association was portable across ancestries at age 9 (African: European: Hispanic), although the association was attenuated among European children at age 15 (African: European: Hispanic). The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: age 15: ), and more accurate than using the top CpG site identified in the metaanalysis, AHRR:CpG05575921 (AUC-ROC age 9:, age 15:, P value).

>>>>

**Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.  DNA methylation signatures in blood and placental samples can function as a biomarker of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues. Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking is associated with salivary polymethylation scores. We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.

**Results**: We created a polymethylation score using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. At age 9, prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. At age 15, At age 9, prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) higher polymetylation score.  The direction and magnitude of the association was portable across genetic ancestries at age 9 although the association was attenuated among European and Hispanic genetic ancestry samples at age 15. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: 0.77 age 15: 0.77), and more accurate than using a single *a priori* CpG site, AHRR:CpG05575921 (AUC-ROC age 9:0.74, *P* value=0.02; age 15:0.73, *P* value=0.02).

### 2021-11-04 Papers/Polymethylation_scores/04_background.Rmd
more revisions post Kelly 1st round comments

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical samples, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation. Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.

An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. <!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa]. Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.

>>>>

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical samples, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation. Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.

An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. Many methods for constructing a DNA methylation biomarker exist, ranging in complexity from single CpG sites to machine learning algorithms such as support vector machines [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. One attractive option is a polymethylation score, analogous to a polygenic score. In a polymethylation score, CpG sites are weighted by the strength of their association with the variable of interest in a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. Using a subset of CpG sites selected with LASSO regression, one such DNA methylation score calculated from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa]. Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at methylation sampling and portable across genetic ancestry groups. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.

### 2021-11-04 Papers/Polymethylation_scores/05_results.Rmd
more revisions post Kelly 1st round comments

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

```{r, echo=F, include=FALSE}
library(knitr)
knit_child(here::here("Reports", "Results.Rmd"))
```

@@

Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique individuals with DNA methylation data measured on the 450K Illumina array, `r m_obs` samples from `r n_total` individuals had data on additional key covariates (Figure 1). `r n_twovisit` individuals had both age 9 and age 15 samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African ancestry and less likely to be from children of Hispanic ancestry (Supplemental Table 1). Twenty percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alchol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic ancestry.

We calculated several common summary measures of DNA methylation, including global methylation, epigenetic clocks (Pediatric and GRIM), and polymethylation scores. We calculated polymethylation scores using regression coefficients for 6074 CpGs made from three different models in Joubert et al.: a regression using newborn cordblood samples controlling for cell-type proportions, a regression using newborn cordblood samples not controlling for cell-type proportions and a regression using peripheral blood samples from older children. We also examined the top probe from Joubert et al., AHRR:cg05575921. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). Notably, polymethylation scores calculated using coefficients from a cell-type corrected regression as weights did not correlate with estimated cell-type proportions, while scores calculated using coefficients from a regression not corrected for cell-type did. Among the `r n_twovisit` individuals with data from both visits, methylation summary measures from the same individual were correlated between the age 9 and age 15 (Supplemental Figure 2). The correlation was strongest for the polymethylation score calculated using weights from a regression for sustained smoking in newborn cordblood with cell-type control (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3; mean pediatric clock age 9: `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))` age 15: `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))`, mean GRIM clock age 9 `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`, age 15: `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3). Based on these results, we focused on the polymethylation score from newborn cordblood with cell-type control, the pediatric clock, AHRR:cg05575921 and global methylation for bivariate and multivariable analyses.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios and were more likely to report prenatal alcohol use, prenatal drug use and postnatal smoking than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European ancestry and had higher prenatal maternal smoking polymethylation scores at both the age 9 and age 15 visit (Table 1). Children of mothers who reported prenatal maternal smoking did not differ from those whose mothers did not in terms of global methylation, epigenetic age estimaated from pediatric clock or estimate cell-type proportions. Compared to children of mothers not reporting prenatal materna smoking, exposed children were significantly more hypomethylated at AHRR:cg05575921 at age 9 (mean methylation `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`) but not age 15 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)` *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

## Multivariable models

The association between prenatal maternal smoking and the cell-type controlled polymethylation score was also observed in multivariable models controlling for precision and confounder variables (Figure 3; Supplemental Table 2). In the entire sample, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 9 and `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 15. A consistent association was observed when stratifying to only African ancestry children (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`). The direction and magnitude of the association was similar in European and Hispanic ancestry children at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2). Prenatal maternal smoking was associated with hypomethylation at AHRR:cg055975921 at age 9 but not age 15 and not associated with global methylation or epigenetic age from the pediatric clock at either age. Results were similar in linear mixed effect models and in sensitivity models controlling for surrogate variables instead of known covariates (Supplemental Tables 3-4; Supplemental Figure 4-5).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A). The polymethylation score with cell-type control had the largest area under the curve (AUC) both when used as a single predictor (age 9 AUC; `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)` ) and when used in models including child sex, maternal income-poverty ratio, immune cell proportion and batch (age 9 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)`). The AUC for the polymethylation score was significantly larger than using only the top CpG from Joubert et al., AHRR:cg05575921 (age 9 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(4)%>%round(3)` , age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(4)%>%round(3)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

>>>>

---
output:
  html_document: default
  word_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

```{r, echo=F, include=FALSE}
library(knitr)
knitr::knit_child(here::here("Reports", "Paper1.Rmd"))
```

@@

Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with DNA methylation data measured on the Illumina 450K  array, `r m_obs` samples from `r n_total` participants had complete data on additional key covariates (Figure 1). There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Hispanic genetic ancestry (Supplemental Table 1). In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic genetic ancestry.

We calculated several summary measures of DNA methylation, including polymethylation scores, global methylation, and epigenetic clocks. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001). Our primary polymethylation score  measure, which was calculated using coefficients from a cell-type corrected regression, was very slightly correlated estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).

We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2). The correlation across ages was strongest for the primary polymethylation score (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3). Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`. Among age 15 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))` and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).

In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from Joubert et al, AHRR:cg05575921, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)` vs `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).  Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=`r round(with(modeldata_people, chisq.test(table(smkPreg_binary, ancestry)))$p.value, 4)`). Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1). At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). At age 15, children exposed to prenatal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

## Multivariable associations between prenatal maternal smoking and DNA methylation summary measures

The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models  (Figure 3; Supplemental Table 2). At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.  At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`). In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` higher polymethylation score. The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).

Prenatal maternal smoking was associated with lower methylation at AHRR:cg055975921 at age 9 (`r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`). Associations between prenatal maternal smoking and AHRR:cg05575921 were less robust across ancestries (Figure 3).

## Sensitivity analyses

When also adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4). The direction of the association between prenatal maternal smoking and AHRR:cg055975921 was also consistent, although no longer significant at age 15. Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.  (Supplemental Figure 5). Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).

Results were similar when using linear mixed effect models instead of age-stratified models (Supplemental Table 3 & 4).

Results were also similar when using polymethylation scores constructed using regression coefficients (see Methods, Supplemental Figure 7). The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.  Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Figure 8).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5). We considered classification when using DNA methylation summary measures alone. We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry. When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.

At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(4)`). Similarly, including the polymethylation significantly improved the base model accuracy at age 15 (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(4)`).

At age 9,the base model with the polymethylation score had a larger AUC than the base model with AHRR:cg05575921 (AUC base model+AHRR:cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`). At age 15, this was also true (AUC for base model+AHRR:cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

## Sensitivity analyses

Results were similar when using other coefficients as the weights in construction of the polymethylation scores. However, the AUC was always significantly higher when using the coefficients from a model which incorporated cell-type control as weights (Supplemental Table 6).
